{\fontsize{14.7}{16}\selectfont Score-Based Tests of Measurement Invariance with Respect to Continuous and Ordinal Variables}

### 2015-12-04 ucl2015/slides.Rnw
presentation materials for seminar at UCL
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@388 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
set.seed(1090)
@

@@

\begin{itemize}
  \item Motivation
  \item Framework
  \item Score-based tests
  \begin{itemize}
    \item Continuous variables
    \item Ordinal variables
    \item Categorial variables
  \end{itemize}
  \item Illustration
\end{itemize}

@@

\textbf{Psychometric models:} Typically measure latent scales based on certain
manifest variables, e.g., item response theory (IRT) models or confirmatory
factor analysis (CFA, today's focus).

@@

\textbf{Problems:}
\begin{itemize}
  \item Subgroups have to be formed in advance.
  \item Continuous variables are often categorized into groups in an ad hoc way
    (e.g., splitting at the median).
  \item In ordinal variables the ordering of the categories is often not exploited
    (assessing only if at least one group differs from the others).
  \item When likelihood ratio or Wald tests are employed, the model has to be
    fitted to each subgroup which can become numerically challenging and computationally
    intensive.
\end{itemize}

@@

\textbf{Model:} Based on log-likelihood $\ell(\cdot)$ for
$p$-dimensional observations $\bm{x}_i$ ($i = 1, \dots, n$)
based on $k$-dimensional parameter ${\bm \theta}$.

@@

\textbf{More generally:} Consider a class of tests that
assesses whether the model ``deviations'' ${\bm s}(\hat{\bm \theta}; \bm{x}_i)$
depend on $v_i$. This can consider only a subset $k*$ of all $k$
parameters/scores or try to capture other patterns than $H_1^*$.

@@

@@

\subsection{Illustration}

\begin{frame}
\frametitle{Illustration}

\textbf{Software:} In R system for statistical computing.
\begin{itemize}
  \item \emph{strucchange} implements this general framwork for
    parameter instability tests.
  \item Object-oriented implementation that can be applied to many
    model classes, including \emph{lavaan} objects for CFA models.
\end{itemize}

@@

\textbf{Data:}
\begin{itemize}
  \item Application of adult gratitude scale to $n=1401$ youth aged 10--19 years.
  \item GQ-6 scale has five Likert scale items with seven points each.
  \item Assess the factor loadings of a one-factor model.
  \item Question: Measurement invariance across six age groups?
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{Illustration}

@@

<<models>>=
gq6cfa <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE,
  group.equal = "loadings")
@

@@

<<wdmo, strip.white=all, fig=TRUE, height=4.5, width=5, include=FALSE>>=
sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
  vcov = "info", functional = "WDMo", plot = TRUE)
@

<<maxlmo, strip.white=all, fig=TRUE, height=4.5, width=5, include=FALSE>>=
sctest(gq6cfa, order.by = yg$agegroup, parm = 1:4,
  vcov = "info", functional = "maxLMo", plot = TRUE)
@

@@

\begin{frame}
\frametitle{Illustration}

@@

@@

\begin{itemize}
  \item General score-based test framework for assessing measurement
    invariance in parametric psychometric models.
  \item Assessment is along some variable $v$ which can be continuous,
    ordinal, or categorical.
  \item Tests can be seen as generalizations of the Lagrange multiplier
    test.
  \item Computation of critical values might require simulation from
    certain stochastic processes (Brownian bridges).
  \item Easy-to-use implementation available in R package \emph{strucchange}.
  \item Can be re-used in model-based recursive partitioning.
\end{itemize}

@@

\small

@@

%% \medskip
%% 
%% Strobl C, Julia Kopf, Zeileis A (2015).
%%   \dquote{Rasch Trees: A New Method for Detecting Differential Item Functioning in the Rasch Model.}
%%   \textit{Psychometrika}. \textbf{80}(2), 289--316.
%%   \doi{10.1007/s11336-013-9388-3}

>>>>

\SweaveOpts{engine=R, eps=FALSE, keep.source=TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("lavaan")
library("strucchange")
library("psychotree")
library("psychotools")
set.seed(1090)
@

@@

\begin{itemize}
  \item Motivation
  \item Framework
  \item Score-based tests
  \begin{itemize}
    \item Continuous variables
    \item Ordinal variables
    \item Categorial variables
  \end{itemize}
  \item Software
  \item Illustrations
\end{itemize}

@@

\textbf{Psychometric models:} Typically measure latent scales based on certain
manifest variables, e.g., item response theory (IRT) models or confirmatory
factor analysis (CFA).

@@

\textbf{Problems:}
\begin{itemize}
  \item Subgroups have to be formed in advance.
  \item Continuous variables are often categorized into groups in an ad hoc way
    (e.g., splitting at the median).
  \item In ordinal variables the ordering of the categories is often not exploited
    -- assessing only if at least one group differs from the others.
  \item When likelihood ratio or Wald tests are employed, the model has to be
    fitted to each subgroup which can become numerically challenging and computationally
    intensive.
\end{itemize}

@@

\textbf{Model:} Based on log-likelihood $\ell(\cdot)$ for
$p$-dimensional observations $\bm{x}_i$ ($i = 1, \dots, n$)
and $k$-dimensional parameter ${\bm \theta}$.

@@

\textbf{More generally:} Consider a class of tests that
assesses whether the model ``deviations'' ${\bm s}(\hat{\bm \theta}; \bm{x}_i)$
depend on $v_i$. This can consider only a subset $k^*$ of all $k$
parameters/scores or try to capture other patterns than $H_1^*$.

@@

<<gefp, echo=FALSE, results=hide>>=
load("mz.rda")
source("../www/mz.R")
gefp_3_info <- gefp(mz_fit, fit = NULL, order.by = mz_data$age, parm = 1:3,
  vcov = info.mzfit, sandwich = FALSE)
gefp_19_info <- gefp(mz_fit, fit = NULL, order.by = mz_data$age,
  vcov = info.mzfit, sandwich = FALSE)
@

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<gefp-dmax, echo=FALSE, results=hide, fig=TRUE, height=4.5, width=9>>=
par(mfcol = c(1, 2))
plot(gefp_3_info,  functional = maxBB, main = "DM, k* = 3", xlab = "Age")
plot(gefp_19_info, functional = maxBB, main = "DM, k* = 19", xlab = "Age")
@

%plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
%plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
%plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
%plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<gefp-cvm, echo=FALSE, results=hide, fig=TRUE, height=4.5, width=9>>=
par(mfcol = c(1, 2))
plot(gefp_3_info,  functional = meanL2BB, main = "CvM, k* = 3", xlab = "Age")
plot(gefp_19_info, functional = meanL2BB, main = "CvM, k* = 19", xlab = "Age")
@

%plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
%plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<gefp-maxlm, echo=FALSE, results=hide, fig=TRUE, height=4.5, width=9>>=
par(mfcol = c(1, 2))
plot(gefp_3_info,  functional = supLM(0.1), main = "max LM, k* = 3", xlab = "Age")
plot(gefp_19_info, functional = supLM(0.1), main = "max LM, k* = 19", xlab = "Age")
@

@@

\subsection{Software}

\begin{frame}
\frametitle{Software}

\textbf{R packages:}
\begin{itemize}
  \item \emph{strucchange} implements this general framwork for
    parameter instability tests.
  \item Object-oriented implementation that can be applied to many
    model classes, including \emph{lavaan} objects for CFA models.
  \item Other psychometric models that cooperate with \emph{strucchange}
    are provided in \emph{psychotools}, e.g., IRT models (Rasch, partial
    credit, rating scale), Bradley-Terry models for paired comparisons,
    and multinomial processing tree models.
  \item Model-based recursive partitioning based on the general
    parameter instability tests are provided in \emph{partykit}.
  \item Adaptation to psychometric models in \emph{psychotree}.
\end{itemize}

\subsection{CFA: Youth gratitude}

\begin{frame}
\frametitle{CFA: Youth gratitude}

\textbf{Question:} Does measurement invariance hold across age groups
when an adult gratitude scale is applied to youth subjects?

@@

\textbf{Source:} Froh \emph{et al.} (2011, \emph{Psychological Assessment}).
  ``Measuring Gratitude in Youth: Assessing the Psychometric Properties of
  Adult Gratitude Scales in Children and Adolescents.''

\textbf{Data:}
\begin{itemize}
  \item GQ-6 gratitude scale with five Likert scale items of seven points each.
  \item Application to $n=1401$ youth aged 10--19 years (six age groups).
  \item Assess the factor loadings of a one-factor model.
\end{itemize}

@@

\begin{frame}[fragile]
\frametitle{CFA: Youth gratitude}

@@

<<models>>=
gq6_cfa <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE,
  group.equal = "loadings")
@

@@

<<wdmo, strip.white=all, fig=TRUE, height=4.5, width=5, include=FALSE>>=
sctest(gq6_cfa, order.by = yg$agegroup, parm = 1:4,
  vcov = "info", functional = "WDMo", plot = TRUE)
@

<<maxlmo, strip.white=all, fig=TRUE, height=4.5, width=5, include=FALSE>>=
sctest(gq6_cfa, order.by = yg$agegroup, parm = 1:4,
  vcov = "info", functional = "maxLMo", plot = TRUE)
@

@@

\begin{frame}
\frametitle{CFA: Youth gratitude}

@@

\subsection{IRT: Examining exams}

\begin{frame}
\frametitle{IRT: Examining exams}

\textbf{Question:} Does measurement invariance hold in a Rasch model for single-choice exam results?

\textbf{Source:} Mathematics for first-year business and economics students at Universit\"at Innsbruck.
Online tests (conducted in OpenOLAT) and written exams for 500--1,000 students per semester.

\textbf{Data:} Individual results from an end-term exam.
\begin{itemize}
  \item 729 students (out of 941 registered).
  \item 13~single-choice items with five answer alternatives,
    covering the basics of analysis, linear algebra, financial mathematics.
  \item Two groups with partially different item pools (on the same topics). Individual versions of
    items generated via \emph{exams}.
  \item Correctly solved items yield 100\% of associated points. Items without correct solution can
    either be unanswered (0\%) or with an incorrect answer ($-25\%$). Only considered as binary here.
\end{itemize}

\begin{frame}[fragile]
\frametitle{IRT: Examining exams}

<<packages, results=hide>>=
library("psychotools")
library("psychotree")
@

\textbf{Data:} Load, select first group, and exclude extreme scorers.

<<data>>=
load("MathExam.rda")
mex <- subset(MathExam, group == 1 & nsolved > 0 & nsolved < 13)
@

\begin{frame}[fragile]
\frametitle{IRT: Examining exams}

<<solved-plot, eval=FALSE>>=
plot(mex$solved)
@

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<solved-plot-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
<<solved-plot>>
@

\begin{frame}[fragile]
\frametitle{IRT: Examining exams}

<<credits-plot, eval=FALSE>>=
plot(mex$credits)
@

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<credits-plot-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
<<credits-plot>>
@

\begin{frame}[fragile]
\frametitle{IRT: Examining exams}

<<raschmodel-plot, eval=FALSE>>=
mex_rasch <- raschmodel(mex$solved)
plot(mex_rasch, type = "profile")
@

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<raschmodel-plot-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
<<raschmodel-plot>>
@

\begin{frame}[fragile]
\frametitle{IRT: Examining exams}

<<raschmodel-piplot, eval=FALSE>>=
plot(mex_rasch, type = "piplot")
@

\vspace*{-0.4cm}

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<raschmodel-piplot-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
<<raschmodel-piplot>>
@

\begin{frame}[fragile]
\frametitle{IRT: Examining exams}

<<raschmodel-maxlm, strip.white=all, fig=TRUE, height=6.5, width=7.5, include=FALSE>>=
sctest(mex_rasch, order.by = jitter(mex$tests),
  vcov = "info", functional = "maxLM", plot = TRUE)
@

<<raschmodel-maxlmo, strip.white=all, fig=TRUE, height=6.5, width=7.5, include=FALSE>>=
mex$otests <- cut(mex$tests, breaks = c(0, 14:24, 26),
  ordered = TRUE, labels = c("<= 14", 15:24, ">= 25"))
sctest(mex_rasch, order.by = mex$otests,
  vcov = "info", functional = "maxLMo", plot = TRUE)
@

Clear violation of measurement invariance: Students that performed poorly in the
previous online tests have a different item profile.

\begin{frame}
\frametitle{IRT: Examining exams}

\centering
\includegraphics[width=0.8\textwidth]{slides-raschmodel-maxlm}

\begin{frame}
\frametitle{IRT: Examining exams}

\centering
\includegraphics[width=0.8\textwidth]{slides-raschmodel-maxlmo}

\begin{frame}[fragile]
\frametitle{IRT: Examining exams}

<<raschtree>>=
mex_tree <- raschtree(solved ~ otests + attempt + semester + study,
  data = mex, vcov = "info", ordinal = "L2")
@

\vspace*{-0.3cm}

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<raschtree-fig, echo=FALSE, results=hide, fig=TRUE, height=6.2, width=10>>=
plot(mex_tree)
@

\subsection{Paired comparisons: Modeling topmodels}

\begin{frame}[fragile]
\frametitle{Paired comparisons: Modeling topmodels}

\textbf{Question:} Does measurement invariance hold for a Bradley-Terry
preference scaling of attractiveness?

\textbf{Source:} Strobl, Wickelmaier, Zeileis
(2010, \textit{Journal of Educational and Behavioral Statistics}).
``Accounting for Individual Differences in Bradley-Terry Models by Means of Recursive Partitioning.''

\textbf{Data:}
\begin{itemize}
  \item Paired comparisons of attractiveness for
        \emph{Germany's Next Topmodel~2007} finalists:
	Barbara, Anni, Hana, Fiona, Mandy, Anja.
  \item Survey with 192~respondents at Universit{\"a}t T{\"u}bingen.
  \item Available covariates: Gender, age, familiarty with the TV~show.
  \item Familiarity assessed by yes/no questions:
        (1)~Do you recognize the women?/Do you know the show?
	(2)~Did you watch it regularly?
	(3)~Did you watch the final show?/Do you know who won?
\end{itemize}

\begin{frame}[fragile]
\frametitle{Paired comparisons: Modeling topmodels}

\vspace*{-0.7cm}
\begin{center}
\includegraphics[width=0.25\textwidth]{Barbara.jpg} \hspace*{0.1cm}
\includegraphics[width=0.25\textwidth]{Anni.jpg} \hspace*{0.1cm}
\includegraphics[width=0.25\textwidth]{Hana.jpg}

\vspace*{0.1cm}

\includegraphics[width=0.25\textwidth]{Fiona.jpg} \hspace*{0.1cm}
\includegraphics[width=0.25\textwidth]{Mandy.jpg} \hspace*{0.1cm}
\includegraphics[width=0.25\textwidth]{Anja.jpg}
\end{center}

\begin{frame}[fragile]
\frametitle{Paired comparisons: Modeling topmodels}

<<Topmodel-MOB>>=
data("Topmodel2007", package = "psychotree")
tm_mob <- bttree(preference ~ ., data = Topmodel2007,
  minsplit = 5, ref = "Barbara")
@

\vspace*{-0.3cm}

\setkeys{Gin}{width=1.05\textwidth}
\hspace*{-0.4cm}%
<<Topmodel-plot, echo=FALSE, fig=TRUE, height=6.2, width=10>>=
plot(tm_mob, abbreviate = 1, yscale = c(0, 0.5))
@

\begin{frame}
\frametitle{Paired comparisons: Modeling topmodels}

\textbf{Recursively partitioned preferences:} Standardized ranking from Bradley-Terry model.

\begin{tabular}{|r|rrrrrr|} \hline
 & Barbara & Anni & Hana & Fiona & Mandy & Anja \\ \hline
<<Topmodel-table, echo=FALSE, results=tex>>=
tm_tab <- round(worth(tm_mob), digits = 2)
tm_tab <- cbind(rownames(tm_tab), format(tm_tab, nsmall = 2))
cat(paste(apply(tm_tab, 1, paste, collapse = " & "), collapse = "\\\\\n"))
@
\\ \hline
\end{tabular}

@@

\begin{itemize}
  \item General score-based test framework for assessing measurement
    invariance in parametric psychometric models.
  \item Assessment is along some variable $v$ which can be continuous,
    ordinal, or categorical.
  \item Tests can be seen as generalizations of the Lagrange multiplier
    test.
  \item Computation of critical values might require simulation from
    certain stochastic processes (Brownian bridges).
  \item Easy-to-use implementation available in R package \emph{strucchange}.
  \item Can be re-used in model-based recursive partitioning in R packages
    \emph{partykit} and \emph{psychotree}.
\end{itemize}

@@

\footnotesize

@@

Strobl C, Julia Kopf, Zeileis A (2015).
  \dquote{Rasch Trees: A New Method for Detecting Differential Item Functioning in the Rasch Model.}
  \textit{Psychometrika}. \textbf{80}(2), 289--316.
  \doi{10.1007/s11336-013-9388-3}

Strobl C, Wickelmaier F, Zeileis A (2011).
  \dquote{Accounting for Individual Differences in Bradley-Terry Models by Means of Recursive Partitioning.}
  \textit{Journal of Educational and Behavioral Statistics}, \textbf{36}(2), 135--153.
  \doi{10.3102/1076998609359791}

Zeileis A, Hothorn T, Hornik K (2008).
 \dquote{Model-Based Recursive Partitioning.}
  \textit{Journal of Computational and Graphical Statistics},
  \textbf{17}(2), 492--514.
  \doi{10.1198/106186008X319331}
